### 2022-08-09 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@19d988557e6f7228fa28bc84dd766f7a35ae8383 🚀

A variety of measures of the structure of ecological networks have been
used to describe the organization of species interactions in a
biological community (Delmas et al. 2019). These measures provide
valuable information on the functioning of ecosystems and their
responses to environmental change (e.g., Pascual and Dunne 2006; Gómez,
Perfectti, and Jordano 2011). For instance, Bascompte et al. (2003)
suggest that plant--pollinator and seed-disperser networks have a highly
nested structure that can promote species persistence. Another example,
in food webs, shows that a high connectance can promote the robustness
of the system to species lost (Dunne, Williams, and Martinez 2002).
However, despite the growing literature on the ecological implications
of network structure, the association between many of these measures
impedes our ability to fully understand what drives the structure and
behavior of ecological networks. In particular, nestedness and
modularity are strongly associated in ecological networks (Fortuna et
al. 2010), and network connectance has been shown to be an important
driver of many other emerging network properties (Timothée Poisot and
Gravel 2014). In light of these observations, it is difficult to assess
whether attributed effects of given properties are the artifacts of
other, perhaps simpler, measures. It might be necessary to factor out
the effect of other correlated measures before analyzing the structure
of ecological networks.

One way to tackle this issue is first to recognize that food webs and
other ecological networks are constrained systems. In other words, the
space of possible network configurations shrinks as we know more about a
network structure. Consequently, the range of values that a given
network property can take depends on the values of other properties.
Above all, the structure of ecological networks is constrained by the
number of species, or nodes, present. Food webs with high species
richness typically have a lower connectance (MacDonald, Banville, and
Poisot 2020) than smaller networks. This is because the number of
realized interactions in empirical food webs scales slower than the
number of possible species pairs (MacDonald, Banville, and Poisot 2020).
As shown by Timothée Poisot and Gravel (2014), connectance itself can
constrain different aspects of network structure such as the degree
distribution. Other measures, such as the maximum trophic level, can
also constrain the space of feasible networks.

Statistical and mathematical models can help fill many gaps in our
knowledge about species interactions. Predictive and null models are two
complementary types of models that have been developed in network
ecology for this purpose. On one hand, predictive models can partially
alleviate the Eltonian shortfall, which describes our current lack of
knowledge on food webs and other ecological networks (Hortal et al.
2015). A variety of such models have recently been developed using
machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). On the other hand, null models help
us identify potential ecological mechanisms that drive species
interactions. They do so by comparing empirical data with an unbiased
distribution of measures generated using a set of rules that exclude the
mechanism of interest (Fortuna and Bascompte 2006; Delmas et al. 2019).
Both types of models are frequently topological, i.e.~they often predict
the adjacency matrix or specific measures of network structure without
taking into account species' identity. According to Strydom et al.
(2021), these topological models could be used to make better
predictions of pairwise species interactions by constraining the space
of feasible networks.

@@

The principle of maximum entropy (MaxEnt) is a statistical and
topological model that can be used for both of these purposes, i.e.~to
make predictions of network structure and to better understand processes
shaping ecological networks. Food-web properties that can be derived
using MaxEnt are varied and pertain to different elements of the network
(i.e.~at the species (node), the interaction (edge) or the community
(network) levels). Because MaxEnt is a method of finding probability
distributions that are least biased given partial knowledge about a
system, these properties need to be represented probabilistically.
Otherwise, some data transformation might be needed. For example, at the
species level, MaxEnt can be used to predict the distribution of trophic
levels among species, as well as the distribution of species'
vulnerability (number of predators) and generality (number of preys). By
contrast, at the interaction level, predictions can be made on the
distribution of interaction strengths in weighted food webs. At the
community level, it can give us a probability distribution of potential
networks or of specific measures of their emerging structure. The
applicability and performance of this method depend on the ecological
information available and on our capacity to find the set of state
variables that best represent natural systems and to translate them into
appropriate statistical constraints. MaxEnt has been used in a wide
range of disciplines, from thermodynamics to chemistry and biology
(Martyushev and Seleznev 2006). It has also been proven useful in
ecology, e.g.~in species distribution models (Phillips, Anderson, and
Schapire 2006) and macroecological models (Harte et al. 2008; Harte and
Newman 2014). The probability distribution of the variable of interest
with the maximum entropy value is unique and least biased under the set
of constraints used. These constraints are built using state variables,
i.e.~variables that represent the macrostate of the system. Having a
validated maximum entropy model for the system at hand allows us to make
rigorous predictions using a minimal amount of data, as well as helping
us describe the most important factors driving that system.

The principle of maximum entropy is a mathematical method of finding
probability distributions, strongly rooted in statistical mechanics and
information theory (Jaynes 1957a, 1957b; Harremoës and Topsøe 2001).
Starting from a set of constraints given by prior knowledge of a system
(i.e.~what we call state variables), this method helps us find
least-biased probability distributions subject to the constraints. These
probability distributions are guaranteed to be unique given our prior
knowledge and represent the most we can say about a system without
making more assumptions. For example, if the only thing we know about a
biological community is its average number of individuals per species,
the least-biased inference we could make on its species abundance
distribution is the exponential distribution (Frank and Smith 2011;
Harte and Newman 2014). However, this does not imply that this
distribution will be the best fit to empirical data. The challenge is to
find the right set of constraints that would best reproduce
distributions found in nature.

The least-biased probability distribution given the constraints used is
the one with the highest entropy among all probability distributions
that satisfy these constraints. Entropy is a measure of the average
amount of information given by the outcome of a random variable. Many
measures of entropy have been developed in physics (Beck 2009), but only
a fraction of them could be used as an optimization measure with the
principle of maximum entropy. According to Beck (2009) and Khinchin
(2013), a measure of entropy \(H\) should satisfy four properties in the
discrete case: (1) it should be a function of a probability distribution
\(p(n)\) only; (2) it should be maximized when \(p(n)\) is uniform; (3)
it should not be influenced by outcomes with a null probability; and (4)
it should be independent of the order of information acquisition. The
Shannon's entropy (Shannon 1948)

@@

where \(g_i\) is the mathematical formulation of the constraint \(i\)
and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
maximized by setting to \(0\) its partial derivative with respect to
\(p(n)\). In this contribution, we show how this can be done by deriving
the joint degree distribution analytically from the number of species
and the number of interactions in food webs.

\hypertarget{objectives}{%
\subsection{Objectives}\label{objectives}}

Despite its extensive use in graph and network theories (e.g., Park and
Newman 2004; van der Hoorn, Lippner, and Krioukov 2018), MaxEnt has in
comparison been little used in network ecology. Like other real system
networks, food webs and other ecological networks are represented
mathematically as graphs. However, the very nature of food webs
(directed simple graphs frequently having self-loops) makes the
mathematical optimization of maximum entropy graph models more
complicated than with many other types of (non-ecological) networks.
MaxEnt has nevertheless been used to predict the degree distribution of
bipartite ecological networks from the number of species and the number
of interactions (Williams 2011) and to predict interaction strengths
between species pairs using their relative abundances within an optimal
transportation theory regularized with information entropy (Stock,
Poisot, and De Baets 2021). However, to the best of our knowledge,
MaxEnt has never been used to predict food-web structure directly, even
though food webs are among the most documented and widespread ecological
networks.

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. We
then compared our predictions against empirical data and null and
neutral models commonly used in network ecology. The first approach
consists in deriving constrained probability distributions of given
network properties directly. We derived the joint degree distribution
(the joint probability distribution that a species has a given number of
prey and predators in a food web) of maximum entropy using only the
number of species \(S\) and the number of interactions \(L\) as state
variables. Then, we predicted the degree distribution of maximum entropy
directly from the joint degree distribution since the first is the sum
of the marginal distributions of the second (a species' degree is the
sum of its in and out-degrees). Because of the scarcity of empirical
data on the number of interactions in ecological networks, we present a
method to predict \(L\) from \(S\) (Box 1), thus allowing the prediction
of the joint degree distribution from \(S\) solely. In turn, the second
approach consists in finding, using different constraints, the adjacency
matrix of maximum entropy from which network properties can be measured.
To do so, we used a flexible and heuristic approach based on simulated
annealing (an optimization algorithm) to find networks \emph{close} to
maximum entropy. As discussed above, our choice of algorithm stands from
the very nature of food webs (i.e.~simple directed networks allowing
self-loops) that makes the analytical derivation of a maximum entropy
graph model difficult. We first built our type I MaxEnt network model
constrained by the connectance of the network (i.e.~the ratio
\(L/S^2\)). A comparison of this model against empirical data indicated
that connectance alone was not sufficient to predict many aspects of
network structure. For this reason, we built our type II MaxEnt network
model, which instead uses the whole joint degree sequence as a
constraint. Overall, we found that this second model was much better at
predicting food-web structure than the one based on connectance.

@@

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(n = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the
New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Of these
two datasets, \(19\) networks had data on species' relative abundances
that were used in the neutral model (Box 2). These \(19\) networks
constitute our \emph{abundance dataset}, which is a subset of our
complete dataset.

All code and data to reproduce this article are available at the Open
Science Framework. Data cleaning, simulations and analyses were
conducted in Julia v1.5.4.

\hypertarget{analytical-models-measures-of-maximum-entropy}{%
\section{Analytical models: Measures of maximum
entropy}\label{analytical-models-measures-of-maximum-entropy}}

@@

\begin{figure}
\hypertarget{fig:joint_dd}{%
\centering
\includegraphics{figures/joint_degree_dist.png}
\caption{Relative number of predators (\(k_{in}\)) as a function of the
relative number of preys (\(k_{out}\)) for each species in (a) empirical
and (b) maximum entropy joint degree sequences. Empirical networks
include all food webs archived on Mangal, as well as the New-Zealand and
Tuesday lake datasets (our complete dataset). The predicted joint degree
sequences were obtained after sampling one realization of the joint
degree distribution of maximum entropy for each network. (c) Difference
between predicted and empirical values when species are ordered
according to their total degrees. In all panels, each dot corresponds to
a single species in one of the network of our complete
dataset.}\label{fig:joint_dd}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:kl_diverg}{%
\centering
\includegraphics{figures/kl_divergence.png}
\caption{(a) Probability density of KL divergence between in and
out-degree sequences of empirical and maximum entropy joint degree
sequences. (b) Difference between the KL divergence of empirical and
predicted networks as a function of connectance. In both panels,
empirical networks include all food webs archived on Mangal, as well as
the New-Zealand and Tuesday lake datasets (our complete dataset). The
joint degree sequence of simulated networks was obtained after sampling
one realization of the joint degree distribution of maximum entropy for
each network.}\label{fig:kl_diverg}
}
\end{figure}

@@

\hypertarget{heuristic-models-networks-of-maximum-entropy}{%
\section{Heuristic models: Networks of maximum
entropy}\label{heuristic-models-networks-of-maximum-entropy}}

@@

\hypertarget{tbl:measures_all}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_all}Standardized mean differences of
predicted network measures with empirical data for all networks
(\(n = 257\)). Positive (negative) values indicate that the measure is
overestimated (underestimated) on average. Empirical networks are all
food webs archived on Mangal, as well as the New Zealand and Tuesday
lake food webs (our complete dataset). Null 1: Type I null model based
on connectance. MaxEnt 1: Type I MaxEnt network model based on
connectance. Null 2: Type II null model based on the joint degree
sequence. MaxEnt 2: Type II MaxEnt network model based on the joint
degree sequence. \(\rho\): nestedness measured by the spectral radius of
the adjacency matrix. \emph{maxtl}: maximum trophic level. \emph{diam}:
network diameter. \emph{MxSim}: average maximum similarity between
species pairs. \emph{Cannib}: proportion of cannibal species (self
loops). \emph{Omniv}: proportion of omnivorous species. \emph{entropy}:
SVD entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
null 1 & -0.167 & 0.980 & 1.428 & -0.502 & 2.007 & 1.493 &
0.056\tabularnewline
MaxEnt 1 & -0.226 & 0.831 & 1.274 & -0.524 & 1.982 & 1.863 &
0.106\tabularnewline
null 2 & 0.160 & -0.125 & 0.016 & 0.007 & 1.078 & 0.559 &
-0.023\tabularnewline
MaxEnt 2 & -0.015 & 0.178 & 0.565 & -0.282 & 0.698 & 0.589 &
0.058\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{tbl:measures_abund}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_abund}Standardized mean differences of
predicted network measures with empirical data for networks having
abundance data (\(n = 19\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks are all New Zealand and Tuesday lake food webs with abundance
data (our abundance dataset). Null 1: Type I null model based on
connectance. MaxEnt 1: Type I MaxEnt network model based on connectance.
Null 2: Type II null model based on the joint degree sequence. MaxEnt 2:
Type II MaxEnt network model based on the joint degree sequence.
\(\rho\): nestedness measured by the spectral radius of the adjacency
matrix. \emph{maxtl}: maximum trophic level. \emph{diam}: network
diameter. \emph{MxSim}: average maximum similarity between species
pairs. \emph{Cannib}: proportion of cannibal species (self loops).
\emph{Omniv}: proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
neutral & 0.367 & -0.090 & 0.027 & 0.266 & 6.870 & 0.576 &
-0.083\tabularnewline
null 1 & -0.134 & 0.950 & 1.919 & -0.369 & 2.077 & 0.614 &
0.068\tabularnewline
MaxEnt 1 & -0.229 & 1.020 & 1.946 & -0.355 & 2.215 & 0.801 &
0.121\tabularnewline
null 2 & 0.128 & -0.115 & -0.135 & 0.157 & 1.444 & 0.029 &
-0.021\tabularnewline
MaxEnt 2 & -0.010 & 0.054 & 0.243 & -0.062 & -0.038 & 0.083 &
0.038\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include all food webs archived on Mangal, as
well as the New-Zealand and Tuesday lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulated
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include all food webs
archived on Mangal, as well as the New-Zealand and Tuesday lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the connectance (type I MaxEnt network model) or the
joint degree sequence (type II MaxEnt network model). The predictions of
the types I and II null models are also plotted. Regression lines are
plotted in each panel. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs_rel}
}
\end{figure}

@@

@@

>>>>

A variety of measures of the structure of ecological networks have been
used to describe the organization of species interactions in a
biological community (Delmas et al. 2019). These measures provide
valuable information on the functioning of ecosystems and their
responses to environmental change (e.g., Pascual and Dunne 2006; Gómez,
Perfectti, and Jordano 2011). For instance, Bascompte et al. (2003)
suggest that plant--pollinator and seed-disperser networks have a highly
nested structure that can promote species persistence. Another example,
in food webs, shows that a high connectance can promote the robustness
of the system to species lost (Dunne, Williams, and Martinez 2002).
However, despite the growing literature on the ecological implications
of network structure, the association between many of these measures
impedes our ability to fully understand what drives the structure and
behavior of ecological networks. In particular, nestedness and
modularity are strongly associated in ecological networks (Fortuna et
al. 2010), and network connectance has been shown to be an important
driver of many other emerging network properties (Timothée Poisot and
Gravel 2014). In light of these observations, it is difficult to assess
whether attributed effects of given properties are the artifacts of
other, perhaps simpler, measures.

One way to tackle this issue is first to recognize that food webs and
other ecological networks are constrained systems. In other words, the
space of possible network configurations shrinks as we know more about a
network structure. For example, there is a finite number of networks
with specified numbers of nodes and edges. Indeed, the structure of
ecological networks is first and above all constrained by the number of
species, or nodes, present. Food webs with high species richness
typically have a lower connectance (MacDonald, Banville, and Poisot
2020) than smaller networks. This is because the number of realized
interactions in empirical food webs scales slower than the number of
possible species pairs (MacDonald, Banville, and Poisot 2020). As shown
by Timothée Poisot and Gravel (2014), connectance itself can constrain
different aspects of network structure such as the degree distribution
(i.e.~the probability distribution of the number of interspecific
interactions realized by a species). Other measures, such as the maximum
trophic level, can also constrain the space of feasible networks.

Prior knowledge on the structure of ecological networks is thus
especially useful in the current context of data scarcity about species
interactions (Jordano 2016). Indeed, this Eltonian shortfall (Hortal et
al. 2015) can be partially alleviated using known information about an
ecological network. As suggested by Strydom et al. (2021), network
structure can be used to improve the prediction of pairwise species
interactions when data is lacking by constraining the space of feasible
networks. Similarly, partial knowledge on the structure of an ecological
network can also be used to predict others of its properties by
constraining their range of possible values. This is important given
that many aspects of network structure cannot be measured empirically
without data on pairwise species interactions, a prevailing situation
worldwide (Timothée Poisot et al. 2021).

Understanding the ecological constraints that shape species interactions
networks and predicting their emerging structure are thus two
complementary aims of network ecology. This distinction between
understanding and predicting is essential when using statistical and
mathematical models in network ecology and interpreting them. On one
hand, null models help us identify potential ecological mechanisms that
drive species interactions and constrain ecological networks. Null
models generate a distribution for a target measure using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). The difference between the model and empirical data
helps us evaluate the effect of this ecological process in nature. On
the other hand, predictive models can help fill many gaps on species
interactions data. A variety of such models have recently been developed
using machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). Given the constrained nature of
ecological networks, it is surprising that the principle of maximum
entropy, a rigorous mathematical method designed for both the analysis
and prediction of constrained systems, has been barely used in network
ecology.

@@

The principle of maximum entropy (MaxEnt) is a mathematical method of
finding probability distributions, strongly rooted in statistical
mechanics and information theory (Jaynes 1957a, 1957b; Harremoës and
Topsøe 2001). Starting from a set of constraints given by prior
knowledge of a system (i.e.~what we call state variables), this method
helps us find least-biased probability distributions subject to the
constraints. These probability distributions are guaranteed to be unique
given our prior knowledge and represent the most we can say about a
system without making more assumptions. For example, if the only thing
we know about a biological community is its average number of
individuals per species, the least-biased inference we could make on its
species abundance distribution is the exponential distribution (Frank
and Smith 2011; Harte and Newman 2014). However, this does not imply
that this distribution will be the best fit to empirical data. The
challenge is to find the right set of constraints that would best
reproduce distributions found in nature.

MaxEnt states that the least-biased probability distribution given the
constraints used is the one with the highest entropy among all
probability distributions that satisfy these constraints. Entropy is a
measure of the average amount of information given by the outcome of a
random variable. Many measures of entropy have been developed in physics
(Beck 2009), but only a fraction of them could be used as an
optimization measure with the principle of maximum entropy. According to
Beck (2009) and Khinchin (2013), a measure of entropy \(H\) should
satisfy four properties in the discrete case: (1) it should be a
function of a probability distribution \(p(n)\) only; (2) it should be
maximized when \(p(n)\) is uniform; (3) it should not be influenced by
outcomes with a null probability; and (4) it should be independent of
the order of information acquisition. The Shannon's entropy (Shannon
1948)

@@

where \(g_i\) is the mathematical formulation of the constraint \(i\)
and \(c_i\), its value. Note that \(F\) is just Shannon's entropy to
which we added terms that each sums to zero (\(g_i = c_i\)). \(F\) is
maximized by setting to \(0\) its partial derivative with respect to
\(p(n)\).

The principle of maximum entropy has been used in a wide range of
disciplines, from thermodynamics, chemistry and biology (Martyushev and
Seleznev 2006) to graph and network theory (e.g., Park and Newman 2004;
van der Hoorn, Lippner, and Krioukov 2018). It has also been proven
useful in ecology, e.g.~in species distribution models (Phillips,
Anderson, and Schapire 2006) and macroecological models (Harte et al.
2008; Harte and Newman 2014). In network ecology, MaxEnt has been used
to predict the degree distribution of bipartite networks from the number
of species and the number of interactions (Williams 2011) and to predict
interaction strengths between species pairs using their relative
abundances within an optimal transportation theory regularized with
information entropy (Stock, Poisot, and De Baets 2021). Thus, to the
best of our knowledge, MaxEnt has never been used to predict food-web
structure directly, even though food webs are among the most documented
and widespread ecological networks (Ings et al. 2009).

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of preys). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the set of state variables that best represent natural
systems and to translate them into appropriate statistical constraints.
Having a validated maximum entropy model for the system at hand allows
us to make least-biased predictions using a minimal amount of data, as
well as identify the most important ecological processes shaping that
system.

\hypertarget{analytical-and-heuristic-approaches}{%
\subsection{Analytical and heuristic
approaches}\label{analytical-and-heuristic-approaches}}

In this contribution, we used two complementary approaches to predict
the structure of food webs using the principle of maximum entropy. The
first approach consists in deriving constrained probability
distributions of given network properties analytically, whereas the
second approach consists in finding the adjacency matrix of maximum
entropy heuristically, from which network properties can be measured. We
compared our predictions against empirical data and null and neutral
models commonly used in network ecology. We focus on deterministic and
unweighted (Boolean) food webs in both approaches for data availability
reasons. However, our framework can be applied to all types of
ecological networks and a wide variety of measures.

For the first approach (analytic), we focus on species level properties.
Specifically, we derived the joint degree distribution (i.e.~the joint
probability distribution that a species has a given number of prey and
predators in its network) of maximum entropy using only the number of
species \(S\) and the number of interactions \(L\) as state variables.
Then, we predicted the degree distribution of maximum entropy directly
from the joint degree distribution since the first is the sum of the
marginal distributions of the second. Because of the scarcity of
empirical data on the number of interactions in food webs, we present a
method to predict \(L\) from \(S\) (Box 1), thus allowing the prediction
of the joint degree distribution from \(S\) solely.

For the second approach (heuristic), we focus on network level
properties. We used a flexible and heuristic model based on simulated
annealing (an optimization algorithm) to find the network configuration
\emph{close} to maximum entropy and measured its structure. We developed
this heuristic model because the analytical derivation of a maximum
entropy graph model of food webs is difficult, and because this model is
readily applicable to other types of ecological networks and measures.
Indeed, the mathematical representation of food webs (i.e.~directed
simple graphs frequently having self-loops) makes the optimization of
maximum entropy graph models more complicated than with many other types
of (non-ecological) networks. In other words, deriving a probability
distribution on the graphs themselves is difficult when working with
food webs. We built two types of heuristic MaxEnt models depending on
the constraint used. Our type I MaxEnt model uses the connectance of the
network (i.e.~the ratio \(L/S^2\)) as a constraint, whereas our type II
MaxEnt model uses the whole joint degree sequence as a constraint.

@@

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(n = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the
New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
Lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Our
complete dataset thus contained a total of \(257\) food webs. Of the
New-Zealand and Tuesday Lake datasets, \(19\) networks had data on
species' relative abundances that were used in the neutral model (Box
2). These \(19\) food webs constitute our \emph{abundance dataset},
which is a subset of our complete dataset.

All code and data to reproduce this article are available at the Open
Science Framework. Data cleaning, simulations and analyses were
conducted in Julia v1.6.1.

\hypertarget{analytical-maximum-entropy-models}{%
\section{Analytical maximum entropy
models}\label{analytical-maximum-entropy-models}}

@@

\begin{figure}
\hypertarget{fig:joint_dd}{%
\centering
\includegraphics{figures/joint_degree_dist.png}
\caption{Relative number of predators (\(k_{in}\)) as a function of the
relative number of preys (\(k_{out}\)) for each species in (a) empirical
and (b) maximum entropy joint degree sequences. Empirical networks
include all food webs archived on Mangal, as well as the New-Zealand and
Tuesday Lake datasets (our complete dataset). The predicted joint degree
sequences were obtained after sampling one realization of the joint
degree distribution of maximum entropy for each network. (c) Difference
between predicted and empirical values when species are ordered
according to their total degrees. In all panels, each dot corresponds to
a single species in one of the network of our complete
dataset.}\label{fig:joint_dd}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:kl_diverg}{%
\centering
\includegraphics{figures/kl_divergence.png}
\caption{(a) Probability density of KL divergence between in and
out-degree sequences of empirical and maximum entropy joint degree
sequences. (b) Difference between the KL divergence of empirical and
predicted networks as a function of connectance. In both panels,
empirical networks include all food webs archived on Mangal, as well as
the New-Zealand and Tuesday Lake datasets (our complete dataset). The
joint degree sequence of simulated networks was obtained after sampling
one realization of the joint degree distribution of maximum entropy for
each network.}\label{fig:kl_diverg}
}
\end{figure}

@@

\hypertarget{heuristic-maximum-entropy-models}{%
\section{Heuristic maximum entropy
models}\label{heuristic-maximum-entropy-models}}

@@

\hypertarget{tbl:measures_all}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_all}Standardized mean differences of
predicted network measures with empirical data for all networks
(\(n = 257\)). Positive (negative) values indicate that the measure is
overestimated (underestimated) on average. Empirical networks are all
food webs archived on Mangal, as well as the New Zealand and Tuesday
Lake food webs (our complete dataset). Null 1: Type I null model based
on connectance. MaxEnt 1: Type I MaxEnt network model based on
connectance. Null 2: Type II null model based on the joint degree
sequence. MaxEnt 2: Type II MaxEnt network model based on the joint
degree sequence. \(\rho\): nestedness measured by the spectral radius of
the adjacency matrix. \emph{maxtl}: maximum trophic level. \emph{diam}:
network diameter. \emph{MxSim}: average maximum similarity between
species pairs. \emph{Cannib}: proportion of cannibal species (self
loops). \emph{Omniv}: proportion of omnivorous species. \emph{entropy}:
SVD entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
null 1 & -0.167 & 0.980 & 1.428 & -0.502 & 2.007 & 1.493 &
0.056\tabularnewline
MaxEnt 1 & -0.226 & 0.831 & 1.274 & -0.524 & 1.982 & 1.863 &
0.106\tabularnewline
null 2 & 0.160 & -0.125 & 0.016 & 0.007 & 1.078 & 0.559 &
-0.023\tabularnewline
MaxEnt 2 & -0.015 & 0.178 & 0.565 & -0.282 & 0.698 & 0.589 &
0.058\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{tbl:measures_abund}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_abund}Standardized mean differences of
predicted network measures with empirical data for networks having
abundance data (\(n = 19\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks are all New Zealand and Tuesday Lake food webs with abundance
data (our abundance dataset). Null 1: Type I null model based on
connectance. MaxEnt 1: Type I MaxEnt network model based on connectance.
Null 2: Type II null model based on the joint degree sequence. MaxEnt 2:
Type II MaxEnt network model based on the joint degree sequence.
\(\rho\): nestedness measured by the spectral radius of the adjacency
matrix. \emph{maxtl}: maximum trophic level. \emph{diam}: network
diameter. \emph{MxSim}: average maximum similarity between species
pairs. \emph{Cannib}: proportion of cannibal species (self loops).
\emph{Omniv}: proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
neutral & 0.367 & -0.090 & 0.027 & 0.266 & 6.870 & 0.576 &
-0.083\tabularnewline
null 1 & -0.134 & 0.950 & 1.919 & -0.369 & 2.077 & 0.614 &
0.068\tabularnewline
MaxEnt 1 & -0.229 & 1.020 & 1.946 & -0.355 & 2.215 & 0.801 &
0.121\tabularnewline
null 2 & 0.128 & -0.115 & -0.135 & 0.157 & 1.444 & 0.029 &
-0.021\tabularnewline
MaxEnt 2 & -0.010 & 0.054 & 0.243 & -0.062 & -0.038 & 0.083 &
0.038\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{figure}
\hypertarget{fig:motifs}{%
\centering
\includegraphics{figures/motifs_distribution.png}
\caption{Motifs profile of empirical, maximum entropy, and null food
webs. Empirical networks include all food webs archived on Mangal, as
well as the New-Zealand and Tuesday Lake datasets (our complete
dataset). Maximum entropy networks were estimated using a simulated
annealing algorithm to find the network of maximum SVD entropy while
maintaining the connectance (type I MaxEnt network model) or the joint
degree sequence (type II MaxEnt network model). The predictions of the
types I and II null models are also plotted. Boxplots display the median
proportions of each motif (middle horizontal lines), as well as the
first (bottom horizontal lines) and third (top horizontal lines)
quartiles. Vertical lines encompass all data points that fall within
\(1.5\) times the interquartile range from both quartiles, and dots are
data points that fall outside this range. Motifs names are from Stouffer
et al. (2007).}\label{fig:motifs}
}
\end{figure}

\begin{figure}
\hypertarget{fig:motifs_rel}{%
\centering
\includegraphics{figures/motifs_relations.png}
\caption{Relationship between motifs proportions of empirical, maximum
entropy, and null food webs. Empirical networks include all food webs
archived on Mangal, as well as the New-Zealand and Tuesday Lake datasets
(our complete dataset). Maximum entropy networks were estimated using a
simulated annealing algorithm to find the network of maximum SVD entropy
while maintaining the connectance (type I MaxEnt network model) or the
joint degree sequence (type II MaxEnt network model). The predictions of
the types I and II null models are also plotted. Regression lines are
plotted in each panel. Motifs names are from Stouffer et al.
(2007).}\label{fig:motifs_rel}
}
\end{figure}

@@

\leavevmode\hypertarget{ref-Ings2009RevEco}{}%
Ings, Thomas C., José M. Montoya, Jordi Bascompte, Nico Blüthgen, Lee
Brown, Carsten F. Dormann, François Edwards, et al. 2009. {``Review:
Ecological Networks Beyond Food Webs.''} \emph{Journal of Animal
Ecology} 78 (1): 253--69.
\url{https://doi.org/10.1111/j.1365-2656.2008.01460.x}.

@@

\leavevmode\hypertarget{ref-Jordano2016SamNeta}{}%
Jordano, Pedro. 2016. {``Sampling Networks of Ecological
Interactions.''} \emph{Functional Ecology} 30 (12): 1883--93.
\url{https://doi.org/10.1111/1365-2435.12763}.
